---
title: "SFP Priority Analysis"
format:
  typst:
    margin:
      x: 1cm
      y: 1cm
    toc: true
    toc-depth: 2

execute:
  echo: false
  warning: false
---

```{=typst}
#set page(numbering: none)
```

## Introduction

For each country, we classify risk-related variables into exposure and
vulnerability categories, normalize the values for fair comparison, and combine
them into a single urgency index. This index highlights countries where
agricultural systems face the greatest combined pressure and risk.

Alongside urgency, we also consider two complementary dimensions: need (where
services and support are most required) and capacity (where delivery is most
feasible) for each area of work. Indicators cover agricultural, economic, and
institutional factors, adjusted to account for differences in country size,
rural population, and agricultural GDP. The resulting indices are categorized
into low, moderate, and high levels and displayed on bivariate maps. Together,
these metrics provide a clear picture of where interventions are both urgent and
actionable, helping to guide investment toward areas with the greatest potential
impact.

```{r setup, include = FALSE}
library(nanoparquet)
library(terra)
library(sf)
library(dplyr)
library(ggplot2)

min_max_norm <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

COUNTRY_SUBSET <- TRUE # Subset initial prioritization countries fron Oscar?
a0_sf_simple <- read_sf("data/a0_vect_vsimple.topojson")
initial_countries <- read.csv("data/initial_prioritization.csv")$ISO3
countries <- if (COUNTRY_SUBSET) initial_countries else a0_sf_simple$iso3_code


prettyprint_countries <- compiler::cmpfun(function(iso3c) {
  country_names <- a0_sf_simple$gaul0_name[match(iso3c, a0_sf_simple$iso3_code)]
  fmat_countries <- sprintf("%s (%s)", country_names, iso3c)
  n_countries <- length(iso3c)
  pretty_str <- sprintf(
    "> **Countries** (%d): %s \n",
    n_countries,
    paste0(fmat_countries, collapse = ", ")
  )
  cat(pretty_str)
})
```

## Urgency

```{r}
#| echo: false
#| include: false

# --- Urgency  ---
urgency_a0 <- read_parquet("data/results/urgency.parquet") |>
  subset(boundary_id == "gaul0_code")

urgency_a0$iso3c <- a0_sf_simple$iso3_code[match(
  urgency_a0$gaul0_code,
  a0_sf_simple$gaul0_code
)]

urgency_a0 <- subset(urgency_a0, iso3c %in% countries)

urgency_a0 <- biscale::bi_class(
  urgency_a0,
  vuln_index,
  expo_index,
  style = "quantile",
  dim = 3
)

urgency_ranks <- urgency_a0[c(
  "iso3c",
  "gaul0_code",
  "bi_class",
  "composite_index",
  "vuln_index_norm",
  "expo_index_norm"
)] |>
  mutate(across(
    -c(gaul0_code, iso3c, bi_class),
    ~ rank(-.x), # rank so that 1 is most urgent
    .names = "{.col}_rank"
  )) |>
  arrange(desc(composite_index))
#
# x <- urgency_ranks$vuln_index
# urgency_ranks$vuln_cut <- cut(
#   urgency_ranks$vuln_index,
#   quantile(x[x > 0], na.rm = TRUE),
#   include.lowest = TRUE
# )
# y <- urgency_ranks$expo_index
#

urgency_top20 <- urgency_ranks |>
  subset(
    composite_index_rank <= 20
  )
```

```{r}
#| fig-align: center
#| fig-width: 10
urgency_sf <- merge(a0_sf_simple, urgency_ranks, by = "gaul0_code") |>
  mutate(
    priority_group = factor(
      case_when(
        bi_class %in% c("1-1", "1-2", "1-3") ~ "Low Priority",
        bi_class %in% c("2-1", "2-2") ~ "Moderate Priority",
        bi_class == "2-3" ~ "Emerging Pressure Zones",
        bi_class == "3-1" ~ "Underserved Opportunity Hotspots",
        bi_class == "3-2" ~ "Strained Systems",
        bi_class == "3-3" ~ "High-Impact Priority Hubs"
      ),
      levels = c(
        "Low Priority",
        "Moderate Priority",
        "Emerging Pressure Zones",
        "Underserved Opportunity Hotspots",
        "Strained Systems",
        "High-Impact Priority Hubs"
      )
    )
  )

ggplot() +
  theme_void(base_size = 14) + # Set a minimal theme for the map
  geom_sf(data = urgency_sf, aes(fill = priority_group), color = NA) +
  geom_sf(
    data = st_as_sf(a0_sf_simple),
    color = "black",
    fill = NA,
    size = 0.2
  ) +
  labs(
    fill = "Priority Groups"
  ) +
  scale_fill_manual(
    values = c(
      "Low Priority" = "#D5D6D2",
      "Moderate Priority" = "#A7C080",
      "Emerging Pressure Zones" = "#DFAF8F",
      "Underserved Opportunity Hotspots" = "#E67E80",
      "Strained Systems" = "#B48EAD",
      "High-Impact Priority Hubs" = "#7FBBB3"
    )
  )
```

### Urgency Focus Groups

#### Underserved Opportunity Hotspots

Populations face acute social and economic challenges. A low production and
population density means these groups may be overlooked in support and
investment, but also may offer opportunities to address production gaps and
decrease vulnerability. The strategy for these systems is to focus on improving
critical production systems and addressing the vulnerability of the population.

```{r}
#| output: asis
urg_hotspot_3_1 <- subset(urgency_ranks, bi_class == "3-1")
prettyprint_countries(urg_hotspot_3_1$iso3c)
```

#### Strained Systems

These countries face high levels of vulnerability and have a moderate production and population density.

```{r}
#| output: asis
urg_hotspot_3_2 <- subset(urgency_ranks, bi_class == "3-2")
prettyprint_countries(urg_hotspot_3_2$iso3c)
```

#### Emerging Pressure Zones

Areas of moderate vulnerability but high population and production values.

```{r}
#| output: asis
urg_hotspot_2_3 <- subset(urgency_ranks, bi_class == "2-3")
prettyprint_countries(urg_hotspot_2_3$iso3c)
```

#### High-impact Priority Hubs

These countries are easy wins where investments can reach large, highly
vulnerable populations and critical production systems, maximizing both urgency
and impact.

These areas are defined by a high population density and high production level
coupled with high levels of vulnerability.

```{r}
#| output: asis
urg_hotspot_3_3 <- subset(urgency_ranks, bi_class == "3-3")
prettyprint_countries(urg_hotspot_3_3$iso3c)
```

## AOW 1

```{r}
# --- AOW 1 ---
aow1_digi <- read_parquet("data/results/aow1_digi.parquet") # high value = more capacity
aow1_need <- read_parquet("data/results/aow1_need.parquet") # high value = more need

aow1 <- merge(aow1_digi, aow1_need, by = "gaul0_code")[c(
  "gaul0_code",
  "capacity_idx",
  "need_idx"
)]
aow1$iso3c <- a0_sf_simple$iso3_code[match(
  aow1$gaul0_code,
  a0_sf_simple$gaul0_code
)]

aow1 <- subset(aow1, iso3c %in% countries)

aow1 <- biscale::bi_class(
  aow1,
  need_idx,
  capacity_idx,
  style = "quantile",
  dim = 3
)
```

```{r}
#| fig-align: center
#| fig-width: 10
aow1_sf <- merge(a0_sf_simple, aow1, by = "gaul0_code") |>
  mutate(
    priority_group = factor(
      case_when(
        bi_class %in% c("1-1", "1-2") ~ "Low Priority",
        bi_class %in% c("2-1", "2-2") ~ "Moderate Priority",
        bi_class %in% c("2-3", "1-3") ~ "Growth Opportunity Zones",
        bi_class == "3-1" ~ "Challenge Hotspots",
        bi_class == "3-2" ~ "Capacity Focus Zones",
        bi_class == "3-3" ~ "Easy Wins"
      ),
      levels = c(
        "Low Priority",
        "Moderate Priority",
        "Growth Opportunity Zones",
        "Challenge Hotspots",
        "Capacity Focus Zones",
        "Easy Wins"
      )
    )
  )

ggplot() +
  theme_void(base_size = 14) + # Set a minimal theme for the map
  geom_sf(data = aow1_sf, aes(fill = priority_group), color = NA) +
  geom_sf(
    data = st_as_sf(a0_sf_simple),
    color = "black",
    fill = NA,
    size = 0.2
  ) +
  labs(
    fill = "Priority Groups"
  ) +
  scale_fill_manual(
    values = c(
      "Low Priority" = "#D5D6D2",
      "Moderate Priority" = "#A7C080",
      "Growth Opportunity Zones" = "#DFAF8F",
      "Challenge Hotspots" = "#E67E80",
      "Capacity Focus Zones" = "#B48EAD",
      "Easy Wins" = "#7FBBB3"
    )
  )
```

### AOW 1 Focus Groups

#### Challenge Hotspots

Areas with high need for support and investment into CIS and Advisory Services.
These regions have high levels of input inefficiency, large yield gaps, and
production levels. However, these regions also have a low capacity and
infrastructure gaps that must be addressed to maximize impact.

```{r}
#| output: asis
aow1_hotspot_3_1 <- subset(aow1, bi_class == "3-1")
prettyprint_countries(aow1_hotspot_3_1$iso3c)
```

#### Growth Opportunity Zones

These countries have a high capacity and minimal limiting factors.

```{r}
#| output: asis
aow1_hotspot_2_3 <- subset(aow1, bi_class %in% c("2-3", "1-3"))
prettyprint_countries(aow1_hotspot_2_3$iso3c)
```

#### Capacity focus zones

These countries have a high need, and moderate capacity to promote project
delivery.

```{r}
#| output: asis
aow1_hotspot_3_2 <- subset(aow1, bi_class == "3-2")
prettyprint_countries(aow1_hotspot_3_2$iso3c)
```

#### Easy Wins

These countries have a high need for support and investment, and also have a
high capacity for delivery through existing infrastructure, education levels,
and other factors.

```{r}
#| output: asis
aow1_hotspot_3_3 <- subset(aow1, bi_class == "3-3")
prettyprint_countries(aow1_hotspot_3_3$iso3c)
```

## AOW 2

```{r}
# --- AOW 2 ---
aow2_shocks_idx <- read_parquet("data/results/aow2_shocks.parquet")
aow2_resilience_idx <- read_parquet("data/results/aow2_resilience.parquet")

aow2 <- merge(aow2_shocks_idx, aow2_resilience_idx, by = c("gaul0_code"))[c(
  "gaul0_code",
  "shock_idx",
  "resilience_idx"
)]
aow2$iso3c <- a0_sf_simple$iso3_code[match(
  aow2$gaul0_code,
  a0_sf_simple$gaul0_code
)]

aow2 <- subset(aow2, iso3c %in% countries)

aow2 <- biscale::bi_class(
  aow2,
  shock_idx,
  resilience_idx,
  style = "quantile",
  dim = 3
)
```

```{r}
#| fig-align: center
#| fig-width: 10
aow2_sf <- merge(a0_sf_simple, aow2, by = "gaul0_code") |>
  mutate(
    priority_group = factor(
      case_when(
        bi_class %in% c("1-1", "1-2", "1-3") ~ "Low Priority",
        bi_class %in% c("2-2", "2-3") ~ "Moderate Priority",
        bi_class == "3-1" ~ "Priority Shock Hotspots",
        bi_class %in% c("3-2", "2-1") ~ "Areas of Concern",
        bi_class == "3-3" ~ "Moderate Priority"
      ),
      levels = c(
        "Low Priority",
        "Moderate Priority",
        "Areas of Concern",
        "Priority Shock Hotspots"
      )
    )
  )

ggplot() +
  theme_void(base_size = 14) + # Set a minimal theme for the map
  geom_sf(data = aow2_sf, aes(fill = priority_group), color = NA) +
  geom_sf(
    data = st_as_sf(a0_sf_simple),
    color = "black",
    fill = NA,
    size = 0.2
  ) +
  labs(
    fill = "Priority Groups"
  ) +
  scale_fill_manual(
    values = c(
      "Low Priority" = "#D5D6D2",
      "Moderate Priority" = "#A7C080",
      "Priority Shock Hotspots" = "#E67E80",
      "Areas of Concern" = "#B48EAD"
    )
  )
```

### AOW 2 Focus Groups

#### Priority Shock Hotspots

Areas with high shocks and low resilience. These systems have a high risk of
crop loss due to climate related shocks and a low ability to recover from
shocks.

```{r}
#| output: asis
aow2_hotspot_3_1 <- subset(aow2, bi_class == "3-1")
prettyprint_countries(aow2_hotspot_3_1$iso3c)
```

#### Areas of Concern

These countries have high risk of climate shocks and a moderate ability to
recover from shocks.

```{r}
#| output: asis
aow2_hotspot_3_2 <- subset(aow2, bi_class %in% c("3-2", "2-1"))
prettyprint_countries(aow2_hotspot_3_2$iso3c)
```

## AOW 3

```{r}
# --- AOW 3 ---
aow3_constraints_idx <- read_parquet("data/results/aow3_constraints.parquet")
aow3_capacity_idx <- read_parquet("data/results/aow3_capacity.parquet")

aow3 <- merge(aow3_constraints_idx, aow3_capacity_idx, by = c("gaul0_code"))[c(
  "gaul0_code",
  "constraint_idx",
  "capacity_idx"
)]
aow3$iso3c <- a0_sf_simple$iso3_code[match(
  aow3$gaul0_code,
  a0_sf_simple$gaul0_code
)]

aow3 <- subset(aow3, iso3c %in% countries)

aow3 <- biscale::bi_class(
  aow3,
  constraint_idx,
  capacity_idx,
  style = "quantile",
  dim = 3
)
```

```{r}
#| fig-align: center
#| fig-width: 10
aow3_sf <- merge(a0_sf_simple, aow3, by = "gaul0_code") |>
  mutate(
    priority_group = factor(
      case_when(
        bi_class %in% c("1-1", "1-2", "1-3") ~ "Low Priority",
        bi_class %in% c("2-2", "2-3") ~ "Moderate Priority",
        bi_class == "3-1" ~ "Priority Constraint Hotspots",
        bi_class %in% c("3-2", "2-1") ~ "Areas of Concern",
        bi_class == "3-3" ~ "Moderate Priority"
      ),
      levels = c(
        "Low Priority",
        "Moderate Priority",
        "Areas of Concern",
        "Priority Constraint Hotspots"
      )
    )
  )

ggplot() +
  theme_void(base_size = 14) + # Set a minimal theme for the map
  geom_sf(data = aow3_sf, aes(fill = priority_group), color = NA) +
  geom_sf(
    data = st_as_sf(a0_sf_simple),
    color = "black",
    fill = NA,
    size = 0.2
  ) +
  labs(
    fill = "Priority Groups"
  ) +
  scale_fill_manual(
    values = c(
      "Low Priority" = "#D5D6D2",
      "Moderate Priority" = "#A7C080",
      "Priority Constraint Hotspots" = "#E67E80",
      "Areas of Concern" = "#B48EAD"
    )
  )
```

### AOW 3 Focus Groups

#### Priority Constraint Hotspots

Areas with severe natural limitations (e.g., poor soils, low pH) and low
capacity (e.g., limited finance or mechanization) to overcome them. These are
prime targets for targeted interventions and investments.

```{r}
#| output: asis
aow3_hotspot_3_1 <- subset(aow3, bi_class == "3-1")
prettyprint_countries(aow3_hotspot_3_1$iso3c)
```

#### Areas of Concern

Areas with moderate natural constraints but still low capacity to address them.
Support here can prevent moderate challenges from becoming critical.

```{r}
#| output: asis
aow3_hotspot_3_2 <- subset(aow3, bi_class %in% c("3-2", "2-1"))

prettyprint_countries(aow3_hotspot_3_2$iso3c)
```

## Local R&D Capacity

```{r}
# --- Capacity/Spending ---
grape_df <- read_parquet("data/results/grape.parquet")
grape_df$gaul0_code <- a0_sf_simple$gaul0_code[match(
  grape_df$iso3c,
  a0_sf_simple$iso3_code
)]

grape_df <- subset(grape_df, iso3c %in% countries)

grape_df <- biscale::bi_class(
  grape_df,
  hr_per_rpop,
  rd_pct_gdp,
  style = "quantile",
  dim = 3
)
```

```{r}
#| fig-align: center
#| fig-width: 10
grape_sf <- merge(a0_sf_simple, grape_df, by = "gaul0_code") |>
  mutate(
    priority_group = factor(
      case_when(
        bi_class == "1-1" ~ "Priority R&D Hotspots",
        bi_class == "1-3" ~ "Human Resource Gap",
        bi_class == "3-1" ~ "R&D Spending Gap",
        bi_class %in% c("1-2", "2-1") ~ "Low R&D Capacity",
        bi_class %in% c("2-2") ~ "Moderate R&D Capacity",
        bi_class %in% c("3-3", "3-2", "2-3") ~ "High R&D Capacity"
      ),
      levels = c(
        "Priority R&D Hotspots",
        "R&D Spending Gap",
        "Human Resource Gap",
        "Low R&D Capacity",
        "Moderate R&D Capacity",
        "High R&D Capacity"
      )
    )
  )

ggplot() +
  theme_void(base_size = 14) + # Set a minimal theme for the map
  geom_sf(data = grape_sf, aes(fill = priority_group), color = NA) +
  geom_sf(
    data = st_as_sf(a0_sf_simple),
    color = "black",
    fill = NA,
    size = 0.2
  ) +
  labs(
    fill = "Priority Groups"
  ) +
  scale_fill_manual(
    values = c(
      "High R&D Capacity" = "#6F8F5F",
      "Moderate R&D Capacity" = "#A7C080",
      "Low R&D Capacity" = "#D8E2C8",
      "Human Resource Gap" = "#7FBBB3",
      "Priority R&D Hotspots" = "#E67E80",
      "R&D Spending Gap" = "#B48EAD"
    )
  )
```

### Focus Groups

#### Priority R&D Hotspots

Areas with low human resources and low R&D spending.

```{r}
#| output: asis
grape_hotspot_1_1 <- subset(grape_df, bi_class == "1-1")
prettyprint_countries(grape_hotspot_1_1$iso3c)
```

#### Human Resource Gap

Areas with low human resources and high R&D spending.

```{r}
#| output: asis
grape_hotspot_1_2 <- subset(grape_df, bi_class == "1-3")
prettyprint_countries(grape_hotspot_1_2$iso3c)
```

#### R&D Spending Gap

Areas with high human resources but a low R&D spending.

```{r}
#| output: asis
grape_hotspot_2_1 <- subset(grape_df, bi_class == "3-1")
prettyprint_countries(grape_hotspot_2_1$iso3c)
```

#### Low R&D Capacity

Areas with low human resources and low R&D spending.

```{r}
#| output: asis
grape_hotspot_1_2 <- subset(grape_df, bi_class %in% c("2-1", "1-2"))
prettyprint_countries(grape_hotspot_1_2$iso3c)
```


{{< pagebreak >}}

## Country Priorities

### Overall Priorities

```{r}
#| fig-align: center
#| fig-width: 10
## Highest Priority Countries

# Urgency
urgency_top_groups <- subset(
  urgency_ranks,
  bi_class %in% c("3-1", "3-1", "3-3", "2-2")
)
# urgency_top_composite <- subset(urgency_ranks, composite_index_rank <= 7)

# R&D
grape_top_groups <- subset(grape_df, bi_class %in% c("1-1"))

# AOW 1
aow1_top_groups <- subset(
  aow1,
  bi_class %in% c("3-1", "3-3", "3-2", "2-3", "1-3")
)

# AOW 2
aow2_top_groups <- subset(aow2, bi_class %in% c("3-2", "2-1", "3-1"))

# AOW 3
aow3_top_groups <- subset(aow3, bi_class %in% c("3-2", "2-1", "3-1"))

occurances <- table(c(
  urgency_top_groups$iso3c,
  # urgency_top_composite$iso3c,
  grape_top_groups$iso3c,
  aow1_top_groups$iso3c,
  aow2_top_groups$iso3c,
  aow3_top_groups$iso3c
))

main_countries <- names(occurances)[occurances > 1]
# Keep highest urgency & remove low
bottom_2s <- names(occurances)[occurances == 2]
top_secondary <- tail(subset(urgency_ranks, iso3c %in% bottom_2s), 2)

# Create Final Groups
top_countries <- main_countries[!(main_countries %in% top_secondary$iso3c)]
secondary_countries <- c(
  top_secondary$iso3c,
  names(occurances)[occurances == 1]
)

priority_plot <- a0_sf_simple |>
  mutate(
    category = case_when(
      iso3_code %in% top_countries ~ "Primary Focus",
      iso3_code %in% secondary_countries ~ "Secondary Focus",
      TRUE ~ "Other"
    )
  )

ggplot() +
  geom_sf(
    data = priority_plot,
    aes(fill = category),
  ) +
  geom_sf(data = a0_sf_simple, color = "black", fill = NA) +
  scale_fill_manual(
    values = c(
      "Primary Focus" = "#E67E80",
      "Secondary Focus" = "#7FBBB3"
    ),
    na.value = NA
  ) +
  labs(
    fill = "Priority Countries"
  ) +
  theme_minimal()
```

#### Primary Focus Countries

```{r}
#| output: asis
prettyprint_countries(top_countries)
```

#### Secondary Focus Countries

```{r}
#| output: asis
prettyprint_countries(secondary_countries)
```

### AOW Priorities

#### AOW 1 Countries

```{r}
#| output: asis

top_aow1 <- subset(aow1, bi_class == "3-3")$iso3c
second_aow1 <- subset(aow1, bi_class %in% c("2-3", "1-3")) # Places with a high capacity

second_aow1$urgency <- urgency_ranks$composite_index[match(
  second_aow1$iso3c,
  urgency_ranks$iso3c
)]

second_aow1 <- second_aow1 |>
  arrange(desc(urgency)) |>
  head(10 - length(top_aow1))


prettyprint_countries(c(top_aow1, second_aow1$iso3c))
```

#### AOW 2 Countries

```{r}
#| output: asis
top_aow2 <- subset(aow2, bi_class == "3-1")$iso3c
second_aow2 <- subset(aow2, bi_class %in% c("2-1", "3-2"))

second_aow2$urgency <- urgency_ranks$composite_index[match(
  second_aow2$iso3c,
  urgency_ranks$iso3c
)]

second_aow2 <- second_aow2 |>
  arrange(desc(urgency)) |>
  head(10 - length(top_aow2))

prettyprint_countries(c(top_aow2, second_aow2$iso3c))
```

#### AOW 3 Countries

```{r}
#| output: asis
subset(aow3, bi_class %in% c("3-2", "2-1", "3-1"))$iso3c |>
  prettyprint_countries()
```

### Country Selection Methodology

The prioritization methodology was based on five dimensions: the Areas of Work
(AOW1–3), urgency, and local research and development capacity. For each
dimension, indicators were compiled into composite indexes representing need,
exposure, constraint, resilience, etc. These which were then used to classify
countries into ranked groups based on quantiles. The top countries were selected
from each of these groups and the countries appearing across multiple dimensions
were counted. The most frequent priority countries across all of these
dimensions were selected as the primary priority countries. The remaining
countries became the secondary priority countries. The top 5 countries for AOW 1
were selected based on high need alongside sufficient infrastructure and human
capacity. The top countries for AOW 2 and AOW 3 were selected as having the the
most shocks or constraining factors paired with the lowest capacity to respond.
Ties for the top 10 countries were broken by the highest urgency index score.

{{< pagebreak >}}

::: {.landscape}

## Appendix 1: Urgency

```{r}
#| fig-align: center
#| fig-width: 20
#| fig-height: 12

urgency_sf |>
  ggplot(aes(x = expo_index_norm * 100, y = vuln_index_norm * 100)) +
  geom_point(aes(col = priority_group)) +
  ggrepel::geom_text_repel(
    aes(label = iso3c),
    size = 5,
    show.legend = FALSE,
    max.overlaps = 100
  ) +
  labs(
    x = "Exposure Index",
    y = "Vulnerability Index",
    title = "Urgency by Country",
    subtitle = "Top 20 countries by urgency index",
    color = "Focus:"
  ) +
  scale_color_manual(
    values = c(
      "Low Priority" = "#D5D6D2",
      "Moderate Priority" = "#A7C080",
      "Emerging Pressure Zones" = "#DFAF8F",
      "Underserved Opportunity Hotspots" = "#E67E80",
      "Strained Systems" = "#B48EAD",
      "High-Impact Priority Hubs" = "#7FBBB3"
    )
  ) +
  theme(legend.position = "bottom")
```

:::
